---
title: "Tobacco Use and Parental Monitoring among American Indian Youth"
author: "Xiaoqi Ma, Avery Turner, Audrey Sileci, & Tracy Zehner"
format: html
editor: visual
execute: 
  eval: true
  echo: false
  error: false
  warning: false
bibliography: references.bib
---

# Abstract

American Indians (AI) and Alaska Natives (AN) report greater cigarette use than all other racial/ethnic groups and subgroups in the U.S. [@Martell2016]. Tobacco use is a major cause of cardiovascular disease (Centers for Disease Control and Prevention, ). In fact, cardiovascular disease and lung cancer are leading causes of death among people who identify as American Indian [@centersfordiseasecontrolandprevention2022]. While past research posits that high rates of tobacco use in AI/AN youth may be due to tobacco companies targeting marginalized groups (FDA, 2023), less is known about how parental monitoring contributes to tobacco use. The purpose of this study was to examine the extent to which parental monitoring predicted tobacco usage and if (AI youth have lower reported parental monitoring) *-- are we able to answer this?*. We hypothesized that higher reported parental monitoring was related to lower tobacco usage rates, on average, in a population of AI and AN adolescents. Data came from the publicly available Substance Use Among American Indian Youth dataset (N =**,** % female, Mage = \_\_ years). We found that parental monitoring\_\_.

# Introduction

People who identify as American Indian (AI) or Alaska Native (AN) report greater cigarette use than all other racial/ethnic groups and subgroups in the U.S. [@martell2016]. In 2022, middle and high-school students who identified as AI/AN were also more likely to report ever having used a tobacco product compared to all other racial/ethnic groups (13.5% of AI/AN compared to 13% who identify as multiracial, 12.4% who identify as white, 11.5% who identify as black, 11.1% who identify as Hispanic/Latinx, and 3.1% who identify as Asian; [@park-lee2022]). Nine out of ten U.S. smokers who report smoking at least daily tried their first cigarette before the age of 18 ***(see note below\*)***. Tobacco use can cause cardiovascular disease and lung cancer which, respectively, are the leading causes of death and cancer deaths among people who identify as AI [@centersfordiseasecontrolandprevention2022]. Identifying risk and protective factors of tobacco use among AI youth is a top priority given the racial disparity in youth smoking rates and the substantial health burden caused by tobacco use in this community.

**\**This link is in Zotero, but won't appear here in R Studio for me: https://www.cdc.gov/tobacco/data_statistics/fact_sheets/youth_data/tobacco_use/index.htm***

Parental monitoring -- parenting behaviors including attention to where children go and activities they engage in [@dishion1998] -- may be a protective factor for decreased substance use among AI youth [@boyd-ball2014; @haruyama2023; @schick2022]. Limited research has been conducted that examines this relation with a focus on tobacco use among AI youth. The challenges that AI parents face are unique. Specifically, traditional AI parenting practices were disrupted by colonization that led to children being separated from their families by forcibly attending boarding schools or being placed with White families, which has left a lasting impact on parental self-efficacy among AI parents [@braveheart1999; @walters2011]. Several support services and interventions are available that target parental monitoring [e.g., @dishion2003], and increasing evidence of parental monitoring and tobacco use among AI youth contributes to the understanding of the potential of these programs in this population.

## The Present Study

Our research questions were the following: (1) Do AI students have lower reported parental monitoring compared to non-AI students? (2) Do AI students with higher reported parental monitoring use substances at lower rates?

(Audrey - justify only AI)

# Methods (Tracy)

## Participants

## Variables of Interest

#### *Independent Variables*

#### *Dependent Variables*

#### *Covariates*

## Analytic Approach

To answer the first research question, we used logistic regression ... in R Version 2023.03.1+446 [@Rcitation]. The packages used for analysis were Tidyverse [@tidy], here [@here], rio [@Chan2018-cy], ggplot2 [@ggplot], missMethods[@missMethods], modelsummary [@modelsummary], janitor [@janitor].

```{r}

library(tidyverse)
library(rio)
library(here)
library(janitor)
library(ggplot2)
library(missMethods)
library(modelsummary)
library(dplyr)

sub <- import(here("data", "37997-0001-Data.rda"))%>%
  clean_names() 

str(sub)
```

# Results

## Descriptive Statistics

```{r}

#Parental monitoring should include all variables that start with b109, including b109_1 (I tell my parents how I'm doing in school), b109_2 (I have to tell my parents who I'm with and what I'm doing at night with friends), b109_3 (My parents ask what I do in my free time), b109_4 (I tell my parents about my activities with friends), b109_5 (I have to tell my parents my plans for weekend nights), b109_6 (My parents ask about things that happen at school), b109_7 (I keep secrets from my parents about what I do in my free time), b109_8 (I need permission to be out late on weeknights), b109_9 (My parents talk to my friends)

#scaling: all questions above are scaled 1-5 (1=never, 2=rarely, 3= sometimes, 4=most of the time, 5= always)

View(sub)

#t-test for RQ1
```

## Parental Monitoring and Tobacco Use

```{r}
#| echo: false
#need to reverse code b109_7

#finding participant identifier
length(unique(sub$enddate, sub$startdate, sub$q2, sub$q1, sub$q3)) 

# TZ: only keep columns of interest (start and end dates will be used for grouping)
R1 <- na.omit(sub %>%
        select(enddate, startdate, b109_1, b109_2, b109_3, b109_4, b109_5, b109_6, b109_7, b109_8, b109_9, q2, q11))
 

# recode to get rid of text in levels; reverse code b109_7
R1_clean <- R1 %>% 
  mutate(b109_1=recode(b109_1, '(1) Never' = "1", '(2) Rarely' = "2", '(3) Sometimes' = "3", '(4) Most of the time' = "4", '(5) Always' = "5")) %>%
  mutate(b109_2=recode(b109_1, '(1) Never' = "1", '(2) Rarely' = "2", '(3) Sometimes' = "3", '(4) Most of the time' = "4", '(5) Always' = "5")) %>%
  mutate(b109_3=recode(b109_3, '(1) Never' = "1", '(2) Rarely' = "2", '(3) Sometimes' = "3", '(4) Most of the time' = "4", '(5) Always' = "5")) %>%
  mutate(b109_4=recode(b109_4, '(1) Never' = "1", '(2) Rarely' = "2", '(3) Sometimes' = "3", '(4) Most of the time' = "4", '(5) Always' = "5")) %>%
  mutate(b109_5=recode(b109_5, '(1) Never' = "1", '(2) Rarely' = "2", '(3) Sometimes' = "3", '(4) Most of the time' = "4", '(5) Always' = "5")) %>%
  mutate(b109_6=recode(b109_6,'(1) Never' = "1", '(2) Rarely' = "2", '(3) Sometimes' = "3", '(4) Most of the time' = "4", '(5) Always' = "5")) %>%
  mutate(b109_7=recode(b109_7, '(1) Never' = "5", '(2) Rarely' = "4", '(3) Sometimes' = "3", '(4) Most of the time' = "2", '(5) Always' = "1")) %>%
  mutate(b109_8=recode(b109_8,'(1) Never' = "1", '(2) Rarely' = "2", '(3) Sometimes' = "3", '(4) Most of the time' = "4", '(5) Always' = "5")) %>%
  mutate(b109_9=recode(b109_9, '(1) Never' = "1", '(2) Rarely' = "2", '(3) Sometimes' = "3", '(4) Most of the time' = "4", '(5) Always' = "5"))

#XM: I thought the only reason we have text in the value is "characterize()" was used when importing data. we could delete that, or use "as.numeric()"?
#XM: For example, I used "sub$b109_7 <- 6-(as.numeric(sub$b109_7))" to reverse code b109_7.
                       
# Creating Factors and Converting Factor to numeric; reverse code b109_7
b109_1 <- factor(R1_clean$b109_1, c(1, 2, 3, 4, 5))
b109_1n<-as.numeric(as.character(b109_1))

b109_2 <- factor(R1_clean$b109_2, c(1, 2, 3, 4, 5))
b109_2n<-as.numeric(as.character(b109_2))

b109_3 <- factor(R1_clean$b109_3, c(1, 2, 3, 4, 5))
b109_3n<-as.numeric(as.character(b109_3))

b109_4 <- factor(R1_clean$b109_4, c(1, 2, 3, 4, 5))
b109_4n<-as.numeric(as.character(b109_4))

b109_5 <- factor(R1_clean$b109_5, c(1, 2, 3, 4, 5))
b109_5n<-as.numeric(as.character(b109_5))

b109_6 <- factor(R1_clean$b109_6, c(1, 2, 3, 4, 5))
b109_6n<-as.numeric(as.character(b109_6))

b109_7 <- factor(R1_clean$b109_7, c(5, 4, 3, 2, 1))
b109_7n<-as.numeric(as.character(b109_7))

b109_8 <- factor(R1_clean$b109_8, c(1, 2, 3, 4, 5))
b109_8n<-as.numeric(as.character(b109_8))

b109_9 <- factor(R1_clean$b109_9, c(1, 2, 3, 4, 5))
b109_9n<-as.numeric(as.character(b109_9))

# binding dataframe and numeric values which looks like: df_new <- cbind(df, new)
R1_clean_num <- cbind(R1_clean, b109_1n, b109_2n, b109_3n, b109_4n, b109_5n, b109_6n, b109_7n, b109_8n, b109_9n)
                       
# create parent monitoring variable from the sum of b109_1:b109_9  (b109_7 was reverse coded) by participant
R1_clean_num <- R1_clean_num%>%
  group_by(startdate, enddate) %>%
  mutate(parent_mon_sum = sum(b109_1n, b109_2n, b109_3n, b109_4n, b109_5n, b109_6n, b109_7n, b109_8n, b109_9n))
##need to standardize parental monitoring var
## maybe use SUMMARIZE instead of mutate                        
                              
         
##Tobacco use Q11 Have you EVER smoked CIGARETTES? (1=never, 2=once or twice, 3=occasionally but not regularly, 4=Regularly in the past, 5=regularly now)

#check which age has the least missingness on our parental monitoring and tobacco use
R1_clean_num <- R1_clean_num %>% rename(ever_smoked = q11, age = q2)
head(R1_clean_num)


Only15yos <- R1_clean_num %>% filter(age=="(15) 15")
table(Only15yos$ever_smoked)
#879
#maybe filter AI & 15yo

#Only16yos <- R1_clean_num %>% filter (age=="(16) 16")
#table(Only16yos$ever_smoked)
#Only17yos <- R1_clean_num %>% filter (age=="(17) 17")
#table(Only17yos$ever_smoked)
#Only18yos <- R1_clean_num %>% filter (age=="(18) 18")
#table(Only18yos$ever_smoked)

#descriptives:_% of 15 yo have ever used cigarettes; parent_mon range and mean; n; sex; gender; race/ethn


#modeling the relationship; filter ai/an
RQ1model <- lm(ever_smoked ~ parent_mon_sum, data=Only15yos)
modelsummary(RQ1model)
```

## AI Students and Parental Monitoring

```{r}
#need to filter by q4_3 (Are you AI?) where 1=yes


```

# Discussion

# References
