---
title: "Substance Use in American Indian Youth"
author: "Xiaoqi Ma, Avery Turner, Audrey Sileci, & Tracy Zehner"
format: html
editor: visual
execute: 
  eval: true
  echo: false
  error: false
  warning: false
bibliography: references.bib
---

# Introduction

## The Present Study

Our research questions were the following: (1) Do students with higher reported parental monitoring use substances at lower rates? (2) Do AI students have lower reported parental monitoring compared to non-AI students?

# Methods

## Participants

## Variables of Interest

#### *Independent Variables*

#### *Dependent Variables*

#### *Covariates*

## Analytic Approach

To answer the first research question, we fit a linear model demonstrating the relation between parental monitoring and substance use using ordinary least squares (OLS) regression in R Version 2023.03.1+446 [@Rcitation]. The packages used for analysis were Tidyverse [@tidy], here [@here], rio [@Chan2018-cy], ggplot2 [@ggplot], and janitor [@janitor].

```{r}

library(tidyverse)
library(rio)
library(here)
library(janitor)
library(ggplot2)

sub <- import(here("data", "37997-0001-Data.rda"))%>%
  clean_names() %>% 
  characterize()
```

# Results

## Descriptive Statistics

```{r}

```

## Parental Monitoring and Substance Use

```{r}
#| echo: false
2 * 2
```

## AI Students and Parental Monitoring

```{r}

```

# Discussion

# References
